---
title: "R Reivew 2023 Lecture 3: Comprehensive Project Workbook"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---


# Set up the R coding Chunks

```{r codechunk_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

0. **Set Working Directory**:
```{r setwd}
#setwd()
```

# Data Exploration and Cleaning:

1. **Generate and Load External Dataset**:
```{r}
set.seed(123)

# Number of clients
n_clients <- 200

# 1. Generate 'name' column
client_data <- data.frame(name = paste0("Client_", seq(1, n_clients)))

# 2. Generate 'car_bought' column. Initially, just a random sample
client_data$car_bought <- sample(rownames(mtcars), n_clients, replace = TRUE)

# 3. Generate 'date_purchased' column
client_data$date_purchased <- sample(seq(as.Date('2015/01/01'), as.Date('2022/01/01'), by="day"), n_clients, replace=TRUE)

# 4. Generate 'income' column
client_data$income <- runif(n_clients, min=30000, max=150000)

# 5. Ensure richer clients are more likely to buy cars with more horsepower
# Sort mtcars by horsepower
sorted_cars <- rownames(mtcars[order(mtcars$hp), ])

# Divide clients into groups and assign cars based on sorted horsepower
split_rows <- ceiling(n_clients / length(sorted_cars))
client_data <- client_data[order(-client_data$income), ] # sort by income
client_data$car_bought <- rep(sorted_cars, times = split_rows)[1:n_clients]

# Show the transformed client_data
head(client_data)

# Write the data to a CSV
write.csv(client_data, "../data/raw/client_data.csv", row.names = FALSE)
```

2. **Load mtcars Dataset and Handle Missing Values**:
```{r}

```

3. **Visualize and Treat Outliers**:
```{r}

```

### Combining Datasets:

4. **Merging the Datasets**:
```{r}

```

### Data Reshaping:

5. **Using `data.table`**:
```{r}

```

### Data Standardization:

6. **Standardizing Date Columns**:
```{r}

```

7. **Standardizing Text Columns**:
```{r}

```

### Unique Identifier Check:

8. **Checking for Unique Identifier**:
```{r}

```

### Save Cleaned Data:

9. **Save Cleaned Data**:
```{r}

```

### Regression Analysis:

10. **Conducting the Regression**:
```{r}

```

11. **Matrix Operations with Coefficients**:
```{r}

```

12. **Loops for Analysis**:
```{r}

```

### Control Structures and Custom Functions:

13. **Using `for` Loop**:
```{r}

```

14. **Using `ifelse`**:
```{r}

```

15. **Using `while` Loop**:
```{r}

```

### The `apply` Family:

16. **Using `lapply`**:
```{r}

```

17. **Using `sapply`**:
```{r}

```


### Advanced Data Manipulation:

19. **Using `tidyverse`**:
```{r}

```

20. **Using `data.table`**:
```{r}

```

## Challenging Question:

21. **Advanced Analysis**:

a. **Calculate Car's Age**:
```{r}

```

b. **Custom Function**:
```{r}

```

c. **Control Structures for Analysis**:
```{r}

```

---
