---
title: "R Reivew 2023 Lecture 3: Comprehensive Project Solution"
author: "Lifeng Ren"
date: "`r Sys.Date()`"
output: html_document
---


# Set up the R coding Chunks

```{r codechunk_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

0. **Set Working Directory**:
```{r setwd}
setwd("~/Library/CloudStorage/Box-Box/Teaching/R_2023/local/lec3/lec3_proj/code")
```

# Data Exploration and Cleaning:

1. **Generate and Load External Dataset**:
```{r}
set.seed(123)

# Number of clients
n_clients <- 200

# 1. Generate 'name' column
client_data <- data.frame(name = paste0("Client_", seq(1, n_clients)))

# 2. Generate 'car_bought' column. Initially, just a random sample
client_data$car_bought <- sample(rownames(mtcars), n_clients, replace = TRUE)

# 3. Generate 'date_purchased' column
client_data$date_purchased <- sample(seq(as.Date('2015/01/01'), as.Date('2022/01/01'), by="day"), n_clients, replace=TRUE)

# 4. Generate 'income' column
client_data$income <- runif(n_clients, min=30000, max=150000)

# 5. Ensure richer clients are more likely to buy cars with more horsepower
# Sort mtcars by horsepower
sorted_cars <- rownames(mtcars[order(mtcars$hp), ])

# Divide clients into groups and assign cars based on sorted horsepower
split_rows <- ceiling(n_clients / length(sorted_cars))
client_data <- client_data[order(-client_data$income), ] # sort by income
client_data$car_bought <- rep(sorted_cars, times = split_rows)[1:n_clients]

# Show the transformed client_data
head(client_data)

# Write the data to a CSV
write.csv(client_data, "../data/raw/client_data.csv", row.names = FALSE)
```

2. **Load mtcars Dataset and Handle Missing Values**:
```{r}
data(mtcars)

mtcars$ID <- rownames(mtcars)

sum(is.na(mtcars))
```

3. **Visualize and Treat Outliers**:
```{r}
boxplot(mtcars$mpg, main="Boxplot of MPG", ylab="Miles Per Gallon")
```

### Combining Datasets:

4. **Merging the Datasets**:
```{r}

# Adjust mtcars to have an ID column matching its row names
mtcars$car_ID <- rownames(mtcars)

# Using base R's merge function
merged_data_base <- merge(mtcars, client_data, by.x="car_ID", by.y="car_bought")

# Using dplyr's join function
library(dplyr)
merged_data_dplyr <- left_join(mtcars, client_data, by=c(car_ID="car_bought"))

```

### Data Reshaping:

5. **Using `data.table`**:
```{r}
library(data.table)
merged_data_table <- as.data.table(merged_data_base)
write.csv(merged_data_table, "../data/temp/merged_data.csv", row.names = FALSE)

# Summary table
summary_table <- merged_data_table[, .(mean_mpg = mean(mpg)), by=car_ID]
print(summary_table)

```

### Data Standardization:

6. **Standardizing Date Columns**:
```{r}
client_data$date_purchased <- as.Date(client_data$date_purchased, format="%Y-%m-%d")
```

7. **Standardizing Text Columns**:
```{r}
library(stringr)
client_data$name <- str_replace(client_data$name, "Client_", "")
```

### Unique Identifier Check:

8. **Checking for Unique Identifier**:
```{r}
length(unique(merged_data_base$name)) == nrow(merged_data_base)

library(eeptools)
isid(merged_data_base, "name", verbose = TRUE)
```

### Save Cleaned Data:

9. **Save Cleaned Data**:
```{r}
write.csv(merged_data_base, "../data/cleaned/merged_data_base_cleaned.csv", row.names = FALSE)
```

### Regression Analysis:

10. **Conducting the Regression**:
```{r}
model <- lm(mpg ~ wt, data=merged_data_base)
plot(merged_data_base$wt, merged_data_base$mpg, main="Relationship between Weight and MPG", 
     xlab="Weight", ylab="Miles Per Gallon", pch=19, col="blue")
abline(model, col="red")
```

11. **Matrix Operations with Coefficients**:
```{r}
coeff_matrix <- matrix(coefficients(model), ncol=1)
identity_matrix <- diag(2)
result_matrix <- identity_matrix %*% coeff_matrix
print(result_matrix)
```

12. **Loops for Analysis**:
```{r}
coefficients_vector <- c()
names_vector <- c()  # To store the names of columns that successfully fit a model

# Filter for rows where 'mpg' is not NA
valid_data <- merged_data_base[!is.na(merged_data_base$mpg), ]

for (column_name in names(valid_data)) {
  if (column_name != "mpg" && is.numeric(valid_data[[column_name]])) {
    # Check for complete cases for the current column and 'mpg'
    complete_rows <- complete.cases(valid_data[, c("mpg", column_name)])
    if (sum(complete_rows) > 0) {
      current_model <- lm(mpg ~ ., data=valid_data[complete_rows, c("mpg", column_name)])
      coefficients_vector <- c(coefficients_vector, coefficients(current_model)[2])
      names_vector <- c(names_vector, column_name)
    }
  }
}

# Plotting the coefficients
barplot(coefficients_vector, names.arg=names_vector, las=2, main="Coefficients of Variables", col="lightblue", cex.names=0.7)

```

### Control Structures and Custom Functions:

13. **Using `for` Loop**:
```{r}
mean_values <- numeric()
for(column in colnames(merged_data_base)){
  if(is.numeric(merged_data_base[[column]])){
    mean_values <- c(mean_values, mean(merged_data_base[[column]], na.rm = TRUE))
  }
}
names(mean_values) <- colnames(merged_data_base)[sapply(merged_data_base, is.numeric)]

print(mean_values)
```

14. **Using `ifelse`**:
```{r}
merged_data_base$mpg_category <- ifelse(merged_data_base$mpg < 20, "Low", 
                                 ifelse(merged_data_base$mpg >= 20 & merged_data_base$mpg < 30, "Medium", "High"))

print(table(merged_data_base$mpg_category))
```

15. **Using `while` Loop**:
```{r}
i <- 1
while(merged_data_base$mpg[i] <= 25){
  i <- i + 1
}
print(paste("The first row where mpg is above 25 is:", i))
```

### The `apply` Family:

16. **Using `lapply`**:
```{r}
range_function <- function(x) c(min=min(x, na.rm=TRUE), max=max(x, na.rm=TRUE))
ranges <- lapply(merged_data_base[, sapply(merged_data_base, is.numeric)], range_function)

print(ranges)
```

17. **Using `sapply`**:
```{r}
column_types <- sapply(merged_data_base, class)
print(column_types)
```


### Advanced Data Manipulation:

19. **Using `tidyverse`**:
```{r}
# Example operations
filtered_data <- filter(merged_data_base, mpg > 20)
selected_data <- select(filtered_data, mpg, wt, gear)
arranged_data <- arrange(selected_data, desc(wt))
```

20. **Using `data.table`**:
```{r}
library(data.table)
merged_data_table[, `:=` (new_mpg = mpg * 1.2)]
```

## Challenging Question:

21. **Advanced Analysis**:

a. **Calculate Car's Age**:
```{r}
# Extracting the year from the purchase date
purchase_year <- as.integer(format(merged_data_base$date_purchased, "%Y"))

# Current year
current_year <- as.integer(format(Sys.Date(), "%Y"))

# Calculate car's age
merged_data_base$car_age <- current_year - purchase_year
```

b. **Custom Function**:
```{r}
stat_function <- function(column){
  c(mean = mean(column, na.rm=TRUE), median = median(column, na.rm=TRUE), sd = sd(column, na.rm=TRUE))
}

numeric_stats <- sapply(merged_data_base[, sapply(merged_data_base, is.numeric)], stat_function)
print(numeric_stats)
```

c. **Control Structures for Analysis**:
```{r}
median_weight <- median(merged_data_base$wt, na.rm=TRUE)
above_median <- merged_data_base[which(merged_data_base$wt > median_weight),]
below_median <- merged_data_base[which(merged_data_base$wt <= median_weight),]

mean_mpg_above_median <- mean(above_median$mpg, na.rm=TRUE)
mean_mpg_below_median <- mean(below_median$mpg, na.rm=TRUE)

print(paste("Average mpg for cars above median weight:", mean_mpg_above_median))
print(paste("Average mpg for cars below or equal to median weight:", mean_mpg_below_median))
```

---
