---
title: "Lecture 3: R-Review-2023"
author: "Lifeng Ren"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: united
    toc: true
    toc_float: true
    toc_depth: 2
    highlight: tango
    df_print: paged
    mathjax: local
    self_contained: false
    number_sections: true
    fig_width: 7
    fig_height: 6
    fig_caption: true
    code_folding: hide
---

# Course Updates

- Lecture 1's link is now available on GitHub


- Any Questions? Comments? Suggestions?


# Recap from Yesterday

- Control structures 
  - Conditional statements (`if`, `else`, `ifelse`)
  - Loops (`for`, `while`, `repeat`, `break`, `next`)
  
- Functions:
  - Introduction to built-in functions
  - Writing custom functions
  - The `apply` family of functions (`lapply`, `sapply`, `tapply`, etc.)
  
- Library and Packages
  
- Data Manipulation
  - `base`
  - `tidyverse`
  - `data.table`

# Introduction

In this lecture, we will delve into the following topics:

- More in Data Cleaning and Wrangling:
  - Revisit the Combining Data sets from yesterday
    - `merge`: `base`
    - `join`: `dplyr`
    - Compare with `1:m`, `m:1`, `1:1` in STATA
  - Data Reshaping
    - `data.table` 
  - Data Standardization
    - Dates: 
    - Text: `stringr`
  - Unique Identifier

- Basic Data Analysis and Data Visualization in R
  - Preliminary Data Checks
  - Simple Regression
  - Results and Visulization
  
- A Comprehensive Project in Class
  - A review for the last 2.5 days
  - Do this in the `rmd` file I provide you.
  - Your output should be a `html` report

---

# For Tomorrow

- Intermediate Econometrics in R Review
  - Mathematical Statistics in R
    - Distributions
    - Simulations
  - Intermediate Applied Econometrics
    - Testing
    - ...

---

# Data Manipulation Again


## Merge: Comparison with STATA

In STATA, when merging datasets, you specify the type of merge using:

* `1:1`: One-to-one join: Each observation in the dataset has a unique identifier, and each identifier in one dataset matches to one and only one identifier in the other dataset.
* `1:m`: One-to-many join: You start with the "many" dataset. For each unique identifier in the "many" dataset, there's a corresponding single observation in the "one" dataset. When you perform the merge, each of the multiple observations in the "many" dataset with the same identifier gets matched to a single observation in the "one" dataset.
* `m:1`: Many-to-one join: You start with the "one" dataset. For each unique identifier in the "one" dataset, there are multiple corresponding observations in the "many" dataset. When you perform the merge, the single observation in the "one" dataset gets matched to each of the multiple observations in the "many" dataset with the same identifier.

---

Let's use examples to understand the link and difference between merging datasets in R and STATA. We will see two cases: 1) all `keys` can be matched and 2) some `keys` cannot be matched:

1. Demo with `R` and save the temporary data into a `.dta` file using the `haven` library.
2. Use `STATA` commands and see if the output matches our expectations.

---

### Every `Key` Exist in Both Data Set

All three should be the same. 

---

#### R Demo using `base`

Since we need to save datasets, we need to define the working directory and libraries we are going to use as usual.

```{r}
setwd("~/Library/CloudStorage/Box-Box/Teaching/R_2023/local/lec3")

library(haven)
```

**One-to-One (`1:1`) Merge:**


```{r}

df1 <- data.frame(ID = c(1, 2, 3), Name = c("Alice", "Bob", "Charlie"))
df2 <- data.frame(ID = c(1, 2, 3), Age = c(25, 30, 28))

write_dta(df1, "df1.dta")
write_dta(df2, "df2.dta")

merged_df <- merge(df1, df2, by = "ID")
write_dta(merged_df, "one_to_one_merge.dta")
```

**Many-to-One (`m:1`) Merge:**

```{r}
many_df <- data.frame(ID = c(1, 1, 2, 3), Score = c(85, 90, 88, 92))
one_df <- data.frame(ID = c(1, 2, 3), Name = c("Alice", "Bob", "Charlie"))

write_dta(many_df, "many_df.dta")
write_dta(one_df, "one_df.dta")

merged_df <- merge(many_df, one_df, by = "ID")
write_dta(merged_df, "many_to_one_merge_R.dta")
```

#### One-to-Many (`1:m`) Merge:

(Using the same example datasets as for the `m:1` merge)

```{r}
merged_df <- merge(one_df, many_df, by = "ID")
write_dta(merged_df, "one_to_many_merge_R.dta")
```

---

#### STATA Demo

We should expect the perfect match.


---

#### R Demo with `dplyr`: `left_join` and `right_join`

For this case, the results should be the same. Because both `many_df` and `one_df` are having information for all IDs. Let us double-check if we are using `left_join`, `right_join` in R. 


First, we'll create two data frames: `teachers` (the "one" dataset) and `classes` (the "many" dataset). Then, we'll perform both m:1 and 1:m merges using both left and right joins, and show the results.

Here's the R code:

```{r}
# Load the required library
library(dplyr)

# Create the 'teachers' dataframe (the "one" dataset)
teachers <- data.frame(
  TeacherID = c(1, 2),
  TeacherName = c("Mr. Smith", "Mrs. Jones")
)

# Create the 'classes' dataframe (the "many" dataset)
classes <- data.frame(
  ClassID = c("A", "B", "C"),
  ClassName = c("Math", "Science", "English"),
  TeacherID = c(1, 1, 2)
)

# m:1 Merge using left join (starting with 'classes' as the base)
m1_merge_left <- left_join(classes, teachers, by = "TeacherID")

# 1:m Merge using left join (starting with 'teachers' as the base)
# This produces the same result as the m:1 merge but potentially with reordered columns
one_m_merge_left <- left_join(teachers, classes, by = "TeacherID")

# m:1 Merge using right join (starting with 'classes' as the base)
m1_merge_right <- right_join(classes, teachers, by = "TeacherID")

# 1:m Merge using right join (starting with 'teachers' as the base)
one_m_merge_right <- right_join(teachers, classes, by = "TeacherID")

# Print results
list(
  m1_merge_left = m1_merge_left,
  one_m_merge_left = one_m_merge_left,
  m1_merge_right = m1_merge_right,
  one_m_merge_right = one_m_merge_right
)
```
Just like before, everything matched!

---

### Some `keys` cannot be matched

Now, let us see the behavior of `left_join` when there's no match for a record in either the left or the right table, and we want this being demonstrated using both `dplyr` and base R. 

Let's slightly modify our datasets to have unmatched records:

1. In the `teachers` dataframe, let's add a teacher with `TeacherID = 3` named "Mr. Doe" who doesn't have any associated class.
2. In the `classes` dataframe, let's add a class with `ClassID = "D"` named "History" with `TeacherID = 4`, but there's no teacher with `TeacherID = 4` in the `teachers` dataframe.

Now, let's perform the left joins using both `dplyr` and base R:

---

#### Using `dplyr`:

1. Perform `left_join` using `classes` as the base.
2. Perform `left_join` using `teachers` as the base.

#### Using base R:

1. Use `merge` with `all.x = TRUE` using `classes` as the base.
2. Use `merge` with `all.x = TRUE` using `teachers` as the base.


```{r}
# Modify the datasets
teachers_modified <- rbind(teachers, data.frame(TeacherID = 3, TeacherName = "Mr. Doe"))
classes_modified <- rbind(classes, data.frame(ClassID = "D", ClassName = "History", TeacherID = 4))

# Using dplyr:
dplyr_classes_base <- left_join(classes_modified, teachers_modified, by = "TeacherID")
dplyr_teachers_base <- left_join(teachers_modified, classes_modified, by = "TeacherID")

# Using base R:
base_classes_base <- merge(classes_modified, teachers_modified, by = "TeacherID", all.x = TRUE)
base_teachers_base <- merge(teachers_modified, classes_modified, by = "TeacherID", all.x = TRUE)

list(
  dplyr_classes_base = dplyr_classes_base,
  dplyr_teachers_base = dplyr_teachers_base,
  base_classes_base = base_classes_base,
  base_teachers_base = base_teachers_base
)
```

**NOTES:**

In both the `dplyr` and base R results, we can observe that where there's no match for a record in the left table, `NA` values are filled in for columns from the right table. Which should be similar to the `m:1` results from STATA's picture below. (NOT EXACT THE SAME BUT SAME IDEA)

```{r echo=FALSE, out.width='50%'}
knitr::include_graphics("https://drive.google.com/uc?export=view&id=1GRP7F4y1kN8w3dU97Y3xVlDIKyoHu1aL")
```

---

## Reshaping


### Base R Methods:

**`reshape` Function:**

Base R provides the `reshape` function, which can convert data from wide to long format and vice versa.

- **Wide to Long**: 

    ```{r}
    # Sample data
    data <- data.frame(
      ID = 1:3,
      Time1 = c(5, 6, 7),
      Time2 = c(8, 9, 10)
    )
    
    print(data)
  
    # Reshaping to long format
    long_data <- reshape(data, direction = "long", varying = list(c("Time1", "Time2")), 
                       v.names = "Value", idvar = "ID", timevar = "Time")
    print(long_data)
    ```

- **Long to Wide**: 

    ```{r}
      # Reshaping to wide format
      wide_data <- reshape(long_data, direction = "wide", v.names = "Value", 
                       idvar = "ID", timevar = "Time")
      print(wide_data)
    ```

### Using `data.table`:

The `data.table` library offers an efficient and flexible approach to data reshaping, especially for large datasets.

**Melting (Wide to Long):**

```{r, message=FALSE}
library(data.table)

# Convert data frame to data table
DT <- as.data.table(data)

# Melt to long format
melted_data <- melt(DT, id.vars = "ID", measure.vars = c("Time1", "Time2"), 
                    variable.name = "Time", value.name = "Value")
print(melted_data)
```

**Casting (Long to Wide):**

```{r}
# Cast to wide format
casted_data <- dcast(melted_data, ID ~ Time, value.var = "Value")
print(casted_data)
```

Remember, the choice between base R and `data.table` methods often depends on your specific needs. `data.table` is especially powerful for large datasets due to its efficiency, while base R can be simpler for basic reshaping tasks or for those who are more familiar with its syntax.

---

## Dates

- Date data can be tricky due to various formats and conventions across the world.
- Proper handling is essential for chronological analysis, time series forecasting, and event tracking.
- `Sys.Date()`: Returns the current date.
- `as.Date()`: Converts a character string into a Date object.
- `format()`: Formats a Date object into a desired character representation.

---

**Examples**
```{r}
# Current date
current_date <- Sys.Date()
print(current_date)

# Convert a string to a date
date_str <- "2023-08-16"
converted_date <- as.Date(date_str)
print(converted_date)

# Format a date
formatted_date <- format(current_date, format="%B %d, %Y")
print(formatted_date)

# Define starting and ending dates
start_date <- as.Date("2023-01-01")
end_date <- as.Date("2023-01-10")

# Create a sequence of dates
date_seq <- seq(start_date, end_date, by="days")

# Convert the sequence to a data frame
date_df <- data.frame(Date = date_seq)

# Print the data frame
print(date_df)

```




---



## Strings

Text data often contains noise in the form of special characters, inconsistencies in formatting, and more. Properly cleaning and manipulating such data is crucial. 

- The `stringr` package in R offers a host of functions that can aid in this. 
- `gsub` in `base` R can do a lot similar functions, I will leave that to you for the future study. Some examples have been provided in the Appendix Section. 

---


### Removing Special Characters

- Special characters can be noise in some analyses.
- Removing them can simplify text and aid in other text processing tasks.
- **`str_replace_all()`**: Replaces all instances of a pattern in a string.

**Examples**
```{r, message=FALSE}
library(stringr)

text <- "Hello, world! This is a test. #Test123"
cleaned_text <- str_replace_all(text, "[^[:alnum:][:space:]]", "")
print(cleaned_text)
```

### String Matching

- Useful to detect if a string contains certain patterns or characters.
- **`str_detect()`**: Detects the presence or absence of a pattern in a string.
- **`str_which()`**: Returns the indices of strings that match a pattern.
- **`str_match()`**: Extract matched groups from a string based on a pattern.

**Examples**
```{r}
# Detect if a string contains "world"
text <- c("Hello world", "Hello R", "R is a world of statistics")
print(str_detect(text, "world"))

# Get indices of strings that contain "R"
print(str_which(text, "R"))

# Extract matched groups
pattern <- "(\\d{4})-(\\d{2})-(\\d{2})"
date_str <- "Today's date is 2023-08-16."
print(str_match(date_str, pattern))
```

### Case Conversion

- To ensure uniformity in text data.
- **`str_to_upper()`**: Converts strings to upper case.
- **`str_to_lower()`**: Converts strings to lower case.
- **`str_to_title()`**: Converts strings to title case.

**Examples**
```{r}
text <- "Hello, World!"

print(str_to_upper(text))
print(str_to_lower(text))
print(str_to_title(text))
```

### Splitting Strings

- To break a string into parts based on a delimiter.
- **`str_split()`**: Splits a string into parts.

**Examples**
```{r}
text <- "apple,banana,grape"
print(str_split(text, ","))
```

### Other `stringr` functions in R with examples
- **`str_length()`**: Computes the length of a string.
- **`str_c()`**: Concatenates strings.
- **`str_sub()`**: Extracts or replaces substrings.

```{r, message=FALSE}
library(stringr)

# String length
print(str_length("Hello, world!"))

# Concatenate
print(str_c("Hello", "world", sep=", "))

# Substring
print(str_sub("Hello, world!", 1, 5))
```

---

## Unique Identifiers in R

- A unique identifier (UID) is an identifier that ensures distinctness among all other items.
- UIDs are crucial for indexing, referencing, and joining datasets without confusion.

### Functions to Identify Unique and Duplicate Values

**`unique()`**: Returns a vector of unique values.
```{r}
vec <- c(1, 2, 2, 3, 4, 4, 4, 5)
unique_vals <- unique(vec)
print(unique_vals)
```

**`duplicated()`**: Returns a logical vector indicating whether an element is a duplicate.
```{r}
vec <- c(1, 2, 2, 3, 4, 4, 4, 5)
dupes <- duplicated(vec)
print(dupes)
```

**`distinct()`** from the `dplyr` package: Used to remove duplicate rows from a data frame or tibble.
```{r}
library(dplyr)
df <- data.frame(name = c("Alice", "Bob", "Alice", "Charlie"), age = c(25, 30, 25, 35))
distinct_df <- distinct(df)
print(distinct_df)
```

### Generating Unique Identifiers

**`seq_along()`**: Generate a sequence along an object's length.
```{r}
vec <- c("apple", "banana", "cherry")
uids <- seq_along(vec)
print(uids)
```

**`make.unique()`**: Generates unique strings by appending numbers.
```{r}
vec <- c("apple", "apple", "banana")
unique_vec <- make.unique(vec)
print(unique_vec)
```

### Checking for Unique Identifiers with `isid()`

The `isid()` function checks if a given set of variables uniquely identifies the observations in a dataset. Just like the one we are using in `STATA`.

- You need to install `eeptools` package for this.

```{r, message=FALSE}
library(eeptools)
df <- data.frame(id = c(1, 2, 3, 1), value = c(10, 20, 30, 40))
# Check if 'id' uniquely identifies the data
isid(df, "id", verbose = TRUE)
```

In this example, the `isid()` function would return `FALSE` because the 'id' variable does not uniquely identify each row in the dataset.

### Considerations

- When working with large datasets or datasets that will be merged, it's essential to ensure that the data's unique identifiers remain consistent.
- Always check for the uniqueness of identifiers, especially before operations like joining or merging datasets, to prevent unintended duplications or omissions.


---

# Basic Data Analysis and Data Visulizations in R

## Preliminary Checks

Before diving into any data analysis, it's essential to understand and clean your data. This involves checking for outliers, handling missing values, and visualizing data distributions.

- For our demonstration, we'll use the `mtcars` dataset.
    ```{r}
    data(mtcars)
    ```

### Checking for Outliers

Outliers can significantly affect regression results. A simple way to check for outliers is by using boxplots.

```{r}
boxplot(mtcars$mpg, main="Boxplot of MPG", ylab="Miles Per Gallon")
```

Points outside the "whiskers" of the boxplot could be potential outliers.

### Handling Missing Values

Data often comes with missing values, and it's crucial to handle them appropriately.

1. **Identifying Missing Values**
   
   Use `is.na()` to identify missing values:
    ```{r}
    missing_vals <- is.na(mtcars$mpg)
    sum(missing_vals)
    ```

2. **Handling Strategies**
   - Remove rows with missing values: `na.omit()`
   - Impute missing values using mean, median, or a specific strategy.

### Data Visualization

Visualizing your data can help in understanding distributions, relationships, and potential issues.

1. **Histogram**: Understand the distribution of a variable.
    ```{r}
    hist(mtcars$mpg, main="Histogram of MPG", xlab="Miles Per Gallon", col="lightblue")
    ```

2. **Correlation Plot**: Understand relationships between variables.
    ```{r}
    pairs(mtcars[, 1:4], main="Scatterplot Matrix")
    ```

3. **Density Plot**: Another way to check the distribution.
    ```{r}
    ggplot(mtcars, aes(x=mpg)) + 
      geom_density(fill="blue", alpha=0.5) +
      labs(title="Density Plot of MPG", x="Miles Per Gallon")
    ```

By conducting these preliminary checks, you ensure that your data is ready for deeper analysis, and any insights or results derived are more likely to be reliable.

---

## Data Analysis and Data Visulization (`mtcars` example)

- Make sure you've installed the packages: `tidyverse`, `lmtest`

- Done the first-round preliminary data check.

### Loading the Libraries

```{r, message=FALSE}
library(tidyverse)
library(lmtest)
```

### Running a Simple Regression in R

A regression allows us to understand relationships between variables. The simplest form is the linear regression, represented as:

\[ Y = \beta_0 + \beta_1 X + \epsilon \]

Where:
- \( Y \) is the dependent variable.
- \( X \) is the independent variable.
- \( \beta_0 \) is the intercept.
- \( \beta_1 \) is the slope.
- \( \epsilon \) is the error term.

```{r}
model <- lm(mpg ~ wt, data = mtcars)
summary(model)
```

The `summary()` function provides a detailed summary of the regression results. Here, we are trying to predict `mpg` (miles per gallon) using the weight (`wt`) of the car, using the `mtcars` dataset.


### Accessing Regression Results

We can save the `summary` of the model first:

```{r}
model_summary <- summary(model)
```

And we can see that it is a `list`! This is something we've already learned.


From this summary object, you can extract:

- `coefficients`: A matrix where each row represents a predictor (including the intercept) and columns provide details like estimate, standard error, t-value, and p-value.
- `sigma`: The residual standard error.
- `r.squared`: The R-squared value.
- `adj.r.squared`: The adjusted R-squared value.
- `fstatistic`: The F-statistic value and its degrees of freedom.

1. **Access the Coefficients**

```{r}
# Access using summary list
coeff_matrix <- model_summary$coefficients
print(coeff_matrix)

# Directly access using `coefficients` functions for regression objects
betas <- coefficients(model) # this will give you only the betas
print(betas)

```

2. **Access the Standard Errors**

The standard errors of the coefficients can be extracted from the model's summary object.

```{r}
standard_errors <- coeff_matrix[, "Std. Error"]
print(standard_errors)
```

3. **Access the t-statistics**

```{r}
t_values <- summary(model)$coefficients[, "t value"]
print(t_values)
```

4. **Access the p-values**

```{r}
p_values <- summary(model)$coefficients[, "Pr(>|t|)"]
print(p_values)
```



## 4. Show Basic Regression Figures

### Scatter Plot with Regression Line

Visualizing the relationship between the independent and dependent variables can be very insightful.

```{r}
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point() +
  geom_smooth(method="lm", col="red") +
  labs(title="Relationship between Car Weight and MPG", x="Weight", y="Miles Per Gallon")
```

### Residuals vs. Fitted Values

Checking the residuals can help diagnose potential issues with the model.

```{r}
residuals <- resid(model)
fitted_values <- fitted(model)

ggplot() +
  geom_point(aes(x=fitted_values, y=residuals)) +
  labs(title="Residuals vs. Fitted Values", x="Fitted Values", y="Residuals")
```

---

# Comprehensive Reivew Project

---

## Introduction

This comprehensive project is designed to review the concepts you've learned throughout the course. You'll apply techniques from data cleaning to advanced data analysis. We'll work with both built-in datasets and an external dataset.

## Datasets

1. **Built-in Dataset**: `mtcars`
2. **External Dataset**: "client_data.csv"

## Workflows

Create a workflow for this project in the following steps:

1. Under **lec3** folder, create a sub folder **lec3_proj**

2. Under **lec3_proj** folder, create two sub folders: **code** and **data**

3. Under **data** folder, create three sub folders: **raw**, **temp**, and **cleaned**

Your directories should look like this:

- **lec3-->**
  - **lec3_proj-->**
    - **code-->**
      - `lec3_proj.Rmd`
    - **data-->**
      - **raw**
      - **temp**
      - **cleaned**

## Tasks

0. **Preparation**
  - Download the `lec3_proj.Rmd` from GitHub into your code folder
  - It is already setup for you with the questions.
  - Code in the code chunk. 
  - Set up your working directory to the folder `~/lec3/lec3_proj/code/`
  


1. **Data Exploration and Cleaning**:
   - Generate an external dataset "client_data.csv" with the following code
   
```{r, eval=FALSE}
set.seed(123)

# Number of clients
n_clients <- 200

# 1. Generate 'name' column
client_data <- data.frame(name = paste0("Client_", seq(1, n_clients)))

# 2. Generate 'car_bought' column. Initially, just a random sample
client_data$car_bought <- sample(rownames(mtcars), n_clients, replace = TRUE)

# 3. Generate 'date_purchased' column
client_data$date_purchased <- sample(seq(as.Date('2015/01/01'), as.Date('2022/01/01'), by="day"), n_clients, replace=TRUE)

# 4. Generate 'income' column
client_data$income <- runif(n_clients, min=30000, max=150000)

# 5. Ensure richer clients are more likely to buy cars with more horsepower
# Sort mtcars by horsepower
sorted_cars <- rownames(mtcars[order(mtcars$hp), ])

# Divide clients into groups and assign cars based on sorted horsepower
split_rows <- ceiling(n_clients / length(sorted_cars))
client_data <- client_data[order(-client_data$income), ] # sort by income
client_data$car_bought <- rep(sorted_cars, times = split_rows)[1:n_clients]

# Show the transformed client_data
head(client_data)

# Write the data to a CSV
write.csv(client_data, "../data/raw/client_data.csv", row.names = FALSE)
```
   
   - Load the built-in dataset `mtcars`.
    - Create a column named `ID`, which is the same as the rownames.
   - Identify missing values using `sum`, and `is.na()` function. 
   - Visualize and treat any outliers in the `mpg` column of the `mtcars` dataset.
    - Hint: using `boxplot()` function, and use `?boxplot()` to learn the syntax.

2. **Combining Datasets**:
   - Merge the datasets using `car_ID`, and `car_bought`.
   - Use both `merge` from `base` and `join` from `dplyr` to combine. Discuss the differences.
   - Note: the merged dataset should be named as:
    - `merged_data_base`
    - `merged_data_dplyr`

3. **Data Reshaping**:
   - Use `data.table` to reshape the merged data set.
   - Save the reshaped data set into the `temp` folder.
   - Create a summary table calculating the mean `mpg` for each unique value in `Car_ID`.

4. **Data Standardization**:
   - Standardize any date columns in "student_data.csv".
    - Purchased Date: `as.Date()`
    - Client's name: get rid of `Client_` using `str_replace()`

5. **Unique Identifier Check**:
   - Check if the mock common column is a unique identifier using:
    - `isid()`
    - `length(unique(DATA$UID))==nrow(DATA)`
   
5.5. **Save and Use Cleaned Data to Proceed**
  - Save the "merged_data_base" into the `cleaned` folder, named as: `merged_data_base_cleaned.csv` 

6. **Regression Analysis**:

   a. **Conducting the Regression**:
      - Using the merged dataset, run a regression predicting `mpg` from another continuous variable.
        - `mpg ~ wt`
      - Visualize the relationship between the chosen predictor and `mpg` using a scatter plot.
        - `plot(merged_data_base$wt, merged_data_base$mpg, ...)`
        - `abline()`

   b. **Matrix Operations with Coefficients**:
      - Extract the coefficients from the regression model and store them in a matrix.
        - new matrix: `matrix()`
        - extract coefficients: `coefficients(model)`
      - Create a 2x2 identity matrix.
        - Use this: `identity_matrix <- diag(2)`
      - Use matrix multiplication (using `%*%`) to multiply the identity matrix by the coefficients matrix. The result will be the coefficients themselves.

   c. **Loops for Analysis (HARD: Don't Do it for Now)**:
      - Write a `for` loop that iterates over the column names of the merged dataset (excluding `mpg`). In each iteration, run a regression using `mpg` as the dependent variable and the current column as the independent variable. Store each coefficient in a vector.
      - After the loop, visualize the coefficients using a bar plot to see the impact of each variable on `mpg`.

   d. **Interpretation**:
      - Discuss the results. Which variables have the most substantial impact on `mpg`? Are the results consistent with expectations?


7. **Control Structures and Custom Functions**:
  - Use a `for` loop to calculate the mean of each numeric column in the merged dataset. (HARD: Don't Do it now). Hints:
    - To calculate the mean of each numeric column in a dataset:
    - Initialization: Start by creating an empty numeric vector to store the mean values.
    - Looping: Use a for loop to iterate over each column in the dataset.
    - Conditional Check: Within the loop, check if the current column is numeric. You can use the is.numeric() function for this.
    - Calculation: If the column is numeric, calculate its mean using the mean() function. Make sure to handle any missing values. Append this mean value to your storage vector.
    - Naming: After the loop completes, name each element in your storage vector with the corresponding column name.
    - Output: Finally, print the named vector to display the mean values.
  - Write a custom function that uses `ifelse` to categorize `mpg` into "Low", "Medium", "High". Apply this function to the dataset.
    - Hint: `ifelse(..., ifelse(,...))`
    
   - Use a `while` loop to find the first row in the merged dataset where `mpg` is above a certain threshold (e.g., 25).


8. **The `apply` Family**:
   - Use `lapply` to calculate the range of each numeric column in the merged dataset. (HARD, don't do it now)
    - hint: `range_function <- function(x) c(min=min(x, na.rm=TRUE), max=max(x, na.rm=TRUE))`
   - Use `sapply` to get the type of each column in the dataset.
    - hint: `sapply(data, class)`

9. **Advanced Data Manipulation**:
   - Use `tidyverse` functions to filter rows, select columns, and arrange the dataset.
    - filter: mpg > 20
    - select: mpg, wt, gear
    - arrange: desc(wt)
   - Use `data.table` to efficiently modify the dataset in place.

## Challenging Question 

10. **Integrative Analysis**:
   - Create a new column in the merged dataset that calculates the age of the car (assumes everyone bought a new car on the purchase_date). Use any necessary libraries/packages.
   - Write a custom function that, given a column name, returns a list containing the mean, median, and standard deviation. Apply this to multiple columns using the `sapply` function.
   - Use control structures to find the average `mpg` for cars that are above and below the median weight. Compare the results and provide an interpretation.

---

# Appendix

## `gsub`

The `gsub()` function is part of base R, and it's a powerful tool for replacing patterns in strings. `gsub()` stands for "global substitution". It searches for all matches of a pattern in a string and replaces them with a specified replacement string.
  
### Syntax

```{r, eval=FALSE}
gsub(pattern, replacement, x, ignore.case = FALSE, perl = FALSE, fixed = FALSE, useBytes = FALSE)
```
- `pattern`: The pattern to search for.
- `replacement`: The string to replace the pattern with.
- `x`: The input string.
- `ignore.case`: Should the match be case-insensitive?
- `perl`: Should Perl-compatible regex be used?
- `fixed`: If TRUE, pattern is a string to be matched as is (turns off special characters).
- `useBytes`: Should bytes be used for matching (relevant for non-ASCII strings)?

### Examples

1. **Basic Substitution**
   Replacing "cat" with "dog":
```{r}
text <- "The cat sat on the mat."
new_text <- gsub("cat", "dog", text)
print(new_text)
```

2. **Removing Special Characters**
   Here's an example using `gsub()` to remove special characters:
```{r}
text <- "Hello, world! This is a test. #Test123"
cleaned_text <- gsub("[^[:alnum:][:space:]]", "", text)
print(cleaned_text)
```

3. **Case-Insensitive Replacement**
   Replacing "world" with "R", ignoring case:
```{r}
text <- "Hello World"
new_text <- gsub("world", "R", text, ignore.case = TRUE)
print(new_text)
```

4. **Replace Multiple Spaces with Single Space**
```{r}
text <- "This   has  multiple    spaces."
new_text <- gsub("\\s+", " ", text)
print(new_text)
```


---

`gsub()` is one of the primary string manipulation functions in base R and is often used in scenarios where you don't want to or can't rely on external packages.

---

# Reference

* Dr. Qingxiao Li's notes for R-Review 2020

* Rodrigo Franco's notes for R-Review 2021









