---
title: "Lecture 5: R-Review-2023"
author: "Lifeng Ren"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: united
    toc: true
    toc_float: true
    toc_depth: 2
    highlight: tango
    df_print: paged
    mathjax: local
    self_contained: false
    number_sections: true
    fig_width: 7
    fig_height: 6
    fig_caption: true
    code_folding: hide
---

# Course Updates

-   This is the last class! Thanks for taking this class!

-   I will reorganize the materials for this class including a solution for today's project and send out an email later. 

-   Any Questions? Comments? Suggestions?

# Recap from Yesterday

- Reviewed Intermediate Mathematical Statistics
  - You should be more comfortable with writing math into R
- Reviewed Regression Analysis
  - You should be comfortable to get access to the data from the summary table of a regression
  
# Introduction

In this lecture, we are going to learn more about `ggplot2` to do data visualization and use another comprehensive project to help you summarize this class with the following topics:

- Download data sets from external website into well-designed working directories
- Load, Append data sets into one data set.
- Data cleaning
  - Re-code 
- Data descriptive statistics:
  - How many observations
  - Unique Identifier
- Data Analysis:
  - Run regression
    - Matrix Calculations with loops by given mathematical expression (e.g.: confidence interval)
  - Data Visualization
- Write, code and output a Rmarkdown report
  
------------------------------------------------------------------------

# Data Visualization with `ggplot2` in R

------------------------------------------------------------------------


## Introduction

`ggplot2` is a powerful and versatile plotting system in R that allows for the creation of complex plots in a structured manner. Today, we'll explore the basics of `ggplot2` and advance to more complex visualizations.

## Basic Visualization with Simulated Data

### Visualize Random Vectors

- Open your `lec5_stu.R` script file
- Generate 100 random numbers from a standard normal distribution use `rnorm(100)`, assign the value the the vector `x`; generate another 100 random numbers and assign it to `error`
  - Remember to `set.seed(ANY NUMBER YOU LIKE)` for reproducible before everything
- Create a vector `y` with this math expression: $y=2x+\epsilon$ 
- Use `plot(x)`, `plot(y)`, `plot(error)`, `plot(x,y)` and check what you see.

```{r}
set.seed(123) # Setting seed for reproducibility

x <- rnorm(100) # Generate 100 random numbers from a standard normal distribution
y <- 2*x + rnorm(100) # Linear relationship with some random noise
error <- rnorm(100) # Error term

# plot(x)
# plot(y)
# plot(error)
plot(x, y)
```

---

**How to use `ggplot2` to do it?**

- Syntax: `⁠ggplot(data = df, mapping = aes(x, y, other aesthetics))⁠`
  - In this case, you need to create a data frame first.
  - Then, you need to define what layer you need:
    - For example, a scatter plot only need `geom_point()` layer
  - You also have the option to choose the theme, use the `theme_minimal()` for now   
```{r}
library(ggplot2)

# Basic scatter plot of x against y
ggplot(data.frame(x=x, y=y), aes(x=x, y=y)) +
  geom_point() +
  theme_minimal()
```

---

**Unknown Distribution**

What if we do not know the distribution and would like to check how each vectors (in the real-world data, it should be variables) is distributed?

```{r}
ggplot(data.frame(x=x), aes(x=x)) +
  geom_density(color="blue", size=1.2) +
  theme_minimal() +
  labs(title="Density of x", x="Value of x", y="Density")
```

### Regression and Visualization

```{r}
# Fit a linear model
model <- lm(y ~ x)

# Plot the data along with the regression line
ggplot(data.frame(x=x, y=y), aes(x=x, y=y)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE, color="blue") + # Regression line
  theme_minimal()
```

---

### Exercise {.tabset}

#### Question

**Question 1**
- Plot the density for error
  - change the color to `red`, size to `0.5`
  - Title change to "Density of Error Term"
  - Label for x-axis: "Value of error terms"
  - Label for y-axis: "Density"
  
---

**Question 2**

- Plot the scatter plot for x and error. Are they Independent by eye-browsing?  
  - For the layers added, try this following code:
  - `geom_point(aes(color = ifelse(x > 0, "blue", "red"))) + scale_color_identity() + theme_minimal()`

---

**Question 3**

- For the regression analysis visualization, copy and paste the code but only change the following line
  - `geom_smooth(method="lm", se=TRUE, color="blue") + # Regression line`
  - What did you see? 

#### Solutions

**Question 1**

```{r}
ggplot(data.frame(x=error), aes(x=error)) +
  geom_density(color="red", size=0.5) +
  theme_minimal() +
  labs(title="Density of Error Term", x="Value of error terms", y="Density")
```
**Question 2**

```{r}
library(ggplot2)

# Basic scatter plot of x against error
ggplot(data.frame(x=x, y=error), aes(x=x, y=error)) +
  geom_point(aes(color = ifelse(x > 0, "blue", "red"))) +
  scale_color_identity() +
  theme_minimal()
```

---

**Question 3**

```{r}
# Fit a linear model
model <- lm(y ~ x)

# Plot the data along with the regression line
ggplot(data.frame(x=x, y=y), aes(x=x, y=y)) +
  geom_point() +
  geom_smooth(method="lm", se=TRUE, color="blue") + # Regression line
  theme_minimal()
```

## 2. Visualization with `mtcars` Dataset

### Basic Histogram

```{r}
# Histogram of mpg (miles per gallon)
ggplot(mtcars, aes(x=mpg)) +
  geom_histogram(binwidth=2, fill="skyblue", color="black") +
  theme_minimal()
```

### Boxplot by Groups

```{r}
# Boxplot of mpg by number of gears
ggplot(mtcars, aes(x=as.factor(gear), y=mpg)) +
  geom_boxplot() +
  theme_minimal()
```

### Faceted Plot

```{r}
# Scatter plot of mpg vs. wt, faceted by number of gears
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(color=as.factor(gear))) +
  facet_wrap(~gear) +
  theme_minimal()
```

---

## Exercise Session: Data Visualization with `mtcars` Dataset using `ggplot2` {.tabset}

### Questions

**Instructions**: Using the `mtcars` dataset and the `ggplot2` library, complete the following exercises. Ensure each plot has an appropriate title and axis labels. If you are unsure how to do a question, use the help function to read the syntax. `?`

---

**Question 1. Modified Histogram**

**Task**: Create a histogram of the `hp` (Horsepower) variable from the `mtcars` dataset. 
- Use a bin width of 10.
- Fill the bars with the color "lightgreen" and set the border color to "darkgreen".

**Hint**: Use the `geom_histogram()` function, similar to the histogram of `mpg`.

---

**2. Boxplot with Modified Aesthetics**

**Task**: Produce a boxplot of `mpg` (miles per gallon) against the number of `carb` (carburetors) in the cars.
- Fill the boxes with the color "lightcoral".
- Set the color of the outliers to "darkred".

**Hint**: Recall the `aes()` function within `geom_boxplot()` to modify outlier aesthetics.

---

**3. Multi-Faceted Scatter Plot**

**Task**: Create a scatter plot that displays the relationship between `drat` (Rear axle ratio) and `qsec` (Quarter mile time). 
- Facet the plot by `cyl` (Number of cylinders) in the cars.
- Color the points based on `vs` (Engine shape: 0 = V-shaped, 1 = straight).
- Ensure different shapes for points based on `vs`.

**Hint**: Use `facet_wrap()` for faceting and consider the `shape` aesthetic in `geom_point()`.

---

### Soluttions

---

**1. Modified Histogram**

```{r}
library(ggplot2)

# Histogram of hp (Horsepower)
ggplot(mtcars, aes(x=hp)) +
  geom_histogram(binwidth=10, fill="lightgreen", color="darkgreen") +
  theme_minimal() +
  labs(title="Histogram of Horsepower", x="Horsepower (hp)", y="Frequency")
```

**Explanation**: 
The `geom_histogram()` function is used to create a histogram of the `hp` column from the `mtcars` dataset. The `fill` and `color` arguments are used to customize the appearance of the histogram bars. The `labs()` function adds a title and axis labels for clarity.

---

**2. Boxplot with Modified Aesthetics**

```{r}
# Boxplot of mpg by number of carburetors
ggplot(mtcars, aes(x=as.factor(carb), y=mpg)) +
  geom_boxplot(fill="lightcoral", outlier.color="darkred") +
  theme_minimal() +
  labs(title="Boxplot of MPG by Number of Carburetors", x="Number of Carburetors", y="Miles per Gallon (mpg)")
```

**Explanation**: 
The `geom_boxplot()` function is used to create a boxplot of the `mpg` column against the number of carburetors (`carb`). The `fill` argument sets the color of the boxes, and the `outlier.color` argument changes the color of the outliers. Again, the `labs()` function is used to provide a title and axis labels.

---

**3. Multi-Faceted Scatter Plot**

```{r}
# Scatter plot of drat vs. qsec, faceted by number of cylinders
ggplot(mtcars, aes(x=drat, y=qsec)) +
  geom_point(aes(color=as.factor(vs), shape=as.factor(vs))) +
  facet_wrap(~cyl) +
  theme_minimal() +
  labs(title="Scatter Plot of Rear Axle Ratio vs. Quarter Mile Time", 
       x="Rear Axle Ratio (drat)", 
       y="Quarter Mile Time (qsec)")
```

**Explanation**: 
The `geom_point()` function creates a scatter plot of `drat` against `qsec`. The points are colored and shaped based on the engine shape (`vs`). The `facet_wrap()` function is used to facet the plot by the number of cylinders (`cyl`). The title and axis labels provide context to the visualization.

---

# Final Comprehensive Project

## Introduction 

For this project you will need to play with the external data source from the website, and as I introduce you at the begining of the class, this project will ask you to finish some tasks to help you review what we have learned in this week. You should finished this project using the `lec5_proj.Rmd` file as your workbook. And remind yourself to create your directories in the following format:

- **lec5-->**
  - **lec5_proj-->**
    - **code-->**
      - `lec5_proj.Rmd`
    - **data-->**
      - **raw**
        - `privateschools2021.xlsx`
      - **temp**
      - **cleaned**

## Tasks and Steps

### Workflows Setup

1. Organize your directories as I showed above. 
2. Create your own `lec5_proj.Rmd` in the `code` folder: just copy and paste what I gave you yesterday to here and change the title.
3. Download data sets from this [website](https://www.cde.ca.gov/ds/si/ps/) into folder `raw`: Only 20-21. Remember that `raw` folder is a subfolder of `data`. Scroll down and download the all the `Excel` files into your `raw` folder. 

### Setup your working directories: `setwd()`

### Load Libraries: 

- `library(readxl)`
- `library(eeptools)`
- `library(dplyr)`
- `library(stringr)`

### Load Data:

4. Use this code to load your data: 
    - `ps2021 <- read_excel("../data/raw/privateschools2021.xlsx", skip = 2)`

### First-round Data Scaning

5. How many observations and variables are there in the data? Use `dim(ps2021)`. In your `Rmd` file, after the code, answer this question. 

6. What is the unique identifier for this data? Hint: test `CDS Code`
    - `isid(df, "id", verbose = TRUE)`
    - `length(unique(DATA$UID))==nrow(DATA)`

### Data Cleaning

7. Use the `library(dplyr)` to select the columns we would like to focus on:
    - County, CDS Code, School Name, Zip, School Phone Number, School Classification, Total Enrollment, Full Time Teachers, Part Time Teachers
    - You need to use the `select` function

8. Use `within()` function to Rename columns, add a new column, and delete a column, here is an example on how to use `within()`

```{r, eval=FALSE}
# Using within() to rename a column and add a new column
data <- within(data, {
  new_column_name <- old_column_name
  new_column <- "New data"
  rm(old_column_name)
})
```

  - Specifically: 
    - Replace the space in the variables' name with underscore.
    - Change all variables' name to lowercase. 
    - Use `tolower()` to change it into lower case
    - Use `str_replace_all` to replace the space. 
  - Try this code inside your `within()` if you cannot figure it out: `names(df_selected) <- str_replace_all(str_to_lower(names(df_selected)), " ", "_")`
  
9. Text Cleaning: get rid of the `-` in the phone number

10. Save the cleaned data into the folder `cleaned`. 

### Data Analysis

11. How many unique School Classification are there in the data? use `length(unique())`

12. Use `ggplot2` with a histogram to show the Distribution of Schools by Religious Classification. 
    - First, you need to goes into school_classification's column and recategorize the classification by changing every values that is not `Non-Religious` or `Other` into `Religious` 
    - Use the histogram to show the distribution for those three larger categories.

13. Run a regression using `total_enrollment` as the dependent variable with `full_time_teachers`

14. Get the regression results' coefficient, and use the coefficients to calculate the confidence interval for 95% confidence level (i.e. z* = 1.96). Suppose you are given the formula to calculate the confidence interval as follows:

$$\beta \pm z^* S E_\beta$$

15. Use `ggplot2` to show the regression results and the confidence interval. 



---

# Reference

-   Dr. Qingxiao Li's notes for R-Review 2020

-   Rodrigo Franco's notes for R-Review 2021
