---
title: "R Reivew 2023 Lecture 3: Comprehensive Project Workbook"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---


```{r codechunk_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

# Setup your working directories: `setwd()`
```{r setwd}
setwd("~/Library/CloudStorage/Box-Box/Teaching/R_2023/repo/R-review2023/lec5/lec5_proj/code")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(dplyr)
library(ggplot2)
library(stringr)
library(eeptools)
library(data.table)
```



### First-round Data Scanning

#### 4. Load Data

```{r}
ps2021 <- read_excel("../data/raw/privateschools2021.xlsx", skip = 2)
head(ps2021)
```


#### 5. Observations and Variables in the Data

```{r}
dim(ps2021)
```

There are 3,050 observations and 53 variables in the data.

#### 6. Unique Identifier for the Data

```{r}
length(unique(ps2021$`CDS Code`)) == nrow(ps2021)
isid(ps2021, "CDS Code", verbose = TRUE)
```

The `CDS Code` is a unique identifier for this dataset.

### Data Cleaning

#### 7. Selecting Specific Columns

```{r}
df_selected <- select(ps2021, County, `CDS Code`, `School Name`, Zip, `School Phone Number`, `School Classification`, `Total Enrollment`, `Full Time Teachers`, `Part Time Teachers`)
head(df_selected)
```

#### 8. Renaming Columns

```{r}
colnames(df_selected) <- str_replace_all(str_to_lower(colnames(df_selected)), " ", "_")

# # Using within():
# 
# df_selected <- within(df_selected, {
#   new_df <- df_selected
#   names(new_df) <- str_replace_all(str_to_lower(names(new_df)), " ", "_")
# })
# 
# head(df_selected)



```

#### 9. Text Cleaning

```{r}
df_selected$school_phone_number <- gsub("-", "", df_selected$school_phone_number)
head(df_selected)
```

#### 10. Saving Cleaned Data

```{r}
write.csv(df_selected, "../data/cleaned/privateschools2021_cleaned.csv")
```

### Data Analysis

#### 11. Unique School Classification

```{r}
length(unique(df_selected$school_classification))
```

There are 34 unique `school_classification` values in the data.

#### 12. Distribution of Schools by Religious Classification

```{r}
df_selected$school_classification <- ifelse(df_selected$school_classification != "Non-Religious" & df_selected$school_classification != "Other", "Religious", df_selected$school_classification)

ggplot(df_selected, aes(x = school_classification)) +
  geom_bar(fill = c("skyblue", "salmon", "lightgreen")) +
  labs(title = "Distribution of Schools by Religious Classification", x = "Classification", y = "Number of Schools")
```

#### 13. Regression Analysis

```{r}
model <- lm(total_enrollment ~ full_time_teachers, data = df_selected)
summary(model)
```

#### 14. Confidence Interval

```{r}
# Run the regression model
model <- lm(total_enrollment ~ full_time_teachers, data = df_selected)

# Extract the coefficient and its standard error
beta <- coef(model)["full_time_teachers"]
se_beta <- summary(model)$coefficients["full_time_teachers", "Std. Error"]

# Calculate the confidence interval
z_star <- 1.96
lower_bound <- beta - z_star * se_beta
upper_bound <- beta + z_star * se_beta

# Display the results
cat("95% Confidence Interval for the coefficient of full_time_teachers:", "\n")
cat("Lower Bound:", lower_bound, "\n")
cat("Upper Bound:", upper_bound, "\n")


confint(model)
```

#### 15. Visualizing Regression Results

```{r}
ggplot(df_selected, aes(x = full_time_teachers, y = total_enrollment)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Regression of Total Enrollment on Full-Time Teachers", x = "Full Time Teachers", y = "Total Enrollment")
```

---

The regression results suggest that for each additional full-time teacher, the total enrollment is expected to increase by about 6.573 students, holding all else constant. The constant (intercept) represents the estimated total enrollment when the number of full-time teachers is zero. The \( R^2 \) value indicates that approximately 42.2% of the variability in total enrollment can be explained by the number of full-time teachers.



