\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[]{amsthm} %lets us use \begin{proof}
\usepackage[]{amssymb} %gives us the character \varnothing
\usepackage{enumerate}
\usepackage{nopageno}
\usepackage{amsfonts}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{times}
\usepackage{amsmath}
\usepackage{physics}
\usepackage{changepage}
\usepackage{graphicx}
\usepackage{subfigure}
\setcounter{MaxMatrixCols}{30}
\usepackage{multirow}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{mdframed}
\usepackage{tcolorbox}
\newcommand{\rd}[1]{{\textcolor{red}{#1}}}
\newcommand{\ph}{\phantom{-}}
\newcommand{\pp}{\phantom{-1}}
\newcommand{\indep}{\perp \!\!\! \perp}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{mathpazo}
\usepackage[framed,numbered,autolinebreaks,useliterate]{mcode}

%\usepackage[colorlinks=true, linkcolor=black, citecolor=blue, urlcolor=blue]{hyperref}
%\usepackage{hyperref}       % This one with red boxes arounded, the above one does not
\usepackage{url}            % simple URL typesetting
\newtheorem{definition}{Definition}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
%\usepackage{lipsum}		% Can be removed after putting your text content
%\usepackage{natbib}
%\usepackage{doi}
%\setcitestyle{aysep={,}}
\newtcolorbox{theo}{colback=red!5!white,colframe=red!75!black,left=5pt,right=5pt,top=2pt,bottom=2pt,}

\newtcolorbox{tip}{colback=blue!5!white,colframe=blue!75!black,left=5pt,right=5pt,top=2pt,bottom=2pt,}
\usepackage{titlesec}
\usepackage{tabularx}
\titleformat*{\section}{\large\bfseries}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

\pagestyle {fancy}
\title{APEC Math Review 2022 Summer}
\author{Instructor: Lifeng Ren}
\date{August 14, 2023 to August 18, 2023}
\lhead{Math Review-Syllabus}
\chead{Lifeng Ren}
\rhead{2023 Summer}


\begin{document}
\thispagestyle{plain}
\begin{center}
{\LARGE \textsc{Introduction to R Statistical Analysis Software}}
\end{center}
\begin{center}
\large \textsc{Syllabus}\\
Summer 2023(August 14, 2023 - August 18, 2023)
\end{center}
%\date{September 26, 2014}

\begin{center}
\rule{6.5in}{0.4pt}
\begin{tabular}{llll}
\textbf{Instructor:} & Lifeng Ren &   \textbf{Time:} & 1PM-4PM, Central Time \\
\textbf{Email:} &\href{mailto:ren00154@umn.edu}{ren00154@umn.edu}  &   \textbf{Place:} & Ruttan 135B and \href{https://umn.zoom.us/j/96218181588?pwd=ZFVrdWQ0bkZyb2N2Wkdwc0JnUk9CQT09}{Zoom}\\
\textbf{Office Hours:} & After class: 4PM-5PM; Ruttan 135B\& \href{https://umn.zoom.us/j/96218181588?pwd=ZFVrdWQ0bkZyb2N2Wkdwc0JnUk9CQT09}{Zoom}&
\end{tabular}
\rule{6.5in}{0.4pt}
\end{center}

\section*{Course Description}
This course is a preliminary introduction to the R statistical software, specifically tailored for new graduate students. All the lecture materials will be provided on the Github Repository (\href{https://github.com/lfr00154/R-review2023}{Here}). Rather than encompassing every aspect of R, the course aims to provide a strong groundwork for further in-depth courses such as Econometric Analysis (APEC8211- 8212), Programming for Econometrics (APEC8221), and Big Data Methods in Economics (APEC8222). Once you're acquainted with R's fundamental aspects, you will be capable of effortlessly managing new functions and packages. If we have time, we will go over some other essential coding and research software and languages such as Quarto, \LaTeX, Git, Markdown, etc.



\section*{Reference}
There are many excellent references out there. Here are some textbooks or online available notes I suggest you guys read or at least have a PDF version in hand for reference. 

\begin{itemize}
    \item \href{https://www.econometrics-with-r.org/index.html}{Introduction to Econometrics with R}
    \item \href{https://libguides.bates.edu/r/r-for-economics}{R for Economics}
    \item \href{https://rkabacoff.github.io/datavis/datavis.pdf}{Data Visualization with R: by Rob Kabacoff}
    \item Matloff, Norman. The art of R programming: A tour of statistical software design. No Starch Press, 2011. (PDF version available through UMN Library or you can find it online)
\end{itemize}
\subsection*{Syntax Cheat Sheet}

\begin{itemize}
    \item \hyperlink{https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf}{Data Wrangling
with dplyr and tidyr}
    \item \href{https://web.mit.edu/hackl/www/lab/turkshop/slides/r-cheatsheet.pdf}{Basic}
    \item \href{https://www.beoptimized.be/pdf/R_Data_Transformation.pdf}{data.table} 
    \item \href{https://nyu-cdsc.github.io/learningr/assets/data-transformation.pdf}{dplyr}
\end{itemize}

\section*{Installation}
\begin{itemize}
    \item Download and Install R and R studio on your desktop from this \href{https://posit.co/download/rstudio-desktop/}{website} 

\end{itemize}

\clearpage
\section*{Exercise}
To learning coding better, we all need practice more. So, we will have several exercises both in-class and after-class. They are \textbf{optional} and are only designed to help you improve your level of understanding. 
\begin{itemize}
    \item I will provide a typed answer key to all exercise I assigned with Quarto's Markdown (.qmd) or R Markdown (.rmd) code (maybe also \LaTeX (.tex) file depending on time). You are free to use my code as a template for yourself.
    \footnote{I strongly recommend you guys spend some time right now to obtain this skill since \LaTeX is a modern way for economists to write papers, present at conferences, etc.}
    \item My teaching goal is to make sure everyone can finish the in-class exercises. If possible, we will spend couple minitues every class to go over the after-class exercise as a review first.   
    
    \end{itemize}


\section*{Class Style}
We will code together and I will expalain the code while we are coding them. We will have a 5 mintues break every hour. 

\section*{Tentative Class Schedule}

\begin{tabularx}{\textwidth}{l|X}
\toprule
Date: & Tasks \\\midrule\\
Before Class: & Install the R, R studio and required libabry\\\\
8/14/2023     &      Workflow; Other necessary coding/study tools for PhD studies. Introduction to R: dataframe and R-studio interface. \\\\
8/15/2023                & Data Manupulation and Data Cleaning: tidyverse, data.table\\\\
8/16/2023          & Functions, loops, and Simulations. \\\\
8/17/2022               &   Econometrics with R:aer \\\\
8/18/2022               & Data Visulization: ggplot2\\
\bottomrule
\end{tabularx}







\end{document}

