---
title: "Lecture 0: R-Review-2023"
author: "Lifeng Ren"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    toc: yes
    toc_float: yes
    toc_depth: 2
    highlight: tango
    df_print: paged
    mathjax: local
    self_contained: no
    number_sections: yes
    fig_width: 7
    fig_height: 6
    fig_caption: yes
    code_folding: hide
  pdf_document:
    toc: yes
    toc_depth: '2'
---
# Set Up

Let us firstly set up our R programming environment and have an overview of Rstudio.

## Install R and Rstudio

Please follow the instruction of this [link](https://posit.co/download/rstudio-desktop/) to download and install R and Rstudio.

## What is R and Rstudio

-   R is a language and environment for statistical computing and graphics.
-   RStudio is an integrated development environment for R.

## Go over the Interface of Rstudio
There are four main windows in Rstudio.

-   The `Console` window

```{r echo=FALSE, out.width='50%'}
knitr::include_graphics("https://drive.google.com/uc?export=view&id=1oCS5ET2HkN95gDVPb286P5Xnx54CM0_3")
```

-   The `Source` window: Here is place we normally write our code. 

```{r echo=FALSE, out.width='50%'}

knitr::include_graphics("https://drive.google.com/uc?export=view&id=1eZqu_8ALnKUPd884ImZaxJI51oaLCShM")
```

-   The `Environment / History / Connections / Tutorial` window: Right now it is empty, because we have not loaded any data yet. Here is the place, we can see some data frames, functions, and vectors. 

```{r echo=FALSE, out.width='50%'}
knitr::include_graphics("https://drive.google.com/uc?export=view&id=1JBkdPzxnAx7jEdzdib95tpkpwvzUSUkd")

```

-   The `Files / Plots / Packages / Help / Viewer` window: You can see your file path, plots, etc. in this window. 

```{r echo=FALSE, out.width='50%'}
knitr::include_graphics("https://drive.google.com/uc?export=view&id=1hSPAOECz8k2Nlae6lgOZyUtF6-iBs5aI")
```



# Code in R

We will talk more about coding in R in the next few days, and there are lot of things to be care of for any programming language, such us: 

* Workflows set up

* Version Control

* Inline Comments

* ...

Each single topic worth a ton of time to study. However, I will only talk about some basics of each part. And I really wish you can code like a pro. 

So, let us get familiar with the `R Script` and `R Markdown` first, and go over some coding conventions.

## R script

You always need yourself and your collaborator know what is your script about. So, please define some basic information for your `R script`. The following is an example I use. 

```{r}

#____________________________
#  Script Information----
#____________________________
##
## Script Title: Introduction to R Statistical Software
##
## Task: Lecture 0
##
## Author: Lifeng Ren
##
## Date Last Modified: 2023-08-14
##
## Date Created: 2023-08-14
##
## Copyright (c) Lifeng Ren, 2023
## Email: ren00154@umn.edu
##
## ___________________________
##
## Version: V1.0 (2023-08-14)
##   
## Version Notes: Initial Efforts
## ___________________________

```

You can create RStudio headers (that can be tracked by Rstudio) using the 

* Windows: `Ctrl + Shift + R`

* Mac OS: `Command + Shift + R`

If you test and try a little bit you would find that R can track and of the following format: `#`+ space + `your section name` + `----` (four dashes).

```{r}
# This is a default R section -------------------------------

#____________________________________________   This is `underscore`
#  Single has tag will not bold the words----   This is `dash`
#____________________________________________   This is `underscore`

#__________________________________________
##  Double hash tag will bold the words----   
#__________________________________________

```




## R Markdown

Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. `.Rmd` is the abbreviation for `R Markdown` file and for more a complete `R markdown` reference, I personally recommend you this [user guide](https://bookdown.org/yihui/rmarkdown/), and official documentation on <http://rmarkdown.rstudio.com> to know the full picture of R markdown. 

- Short note: For the students taking APEC8211 - APEC8212, you need to hand in the Homework in a typed format. So, learning code in `Rmd` would be good to save you time. 

In this class we are going to go over some very basic knowledge of R Markdown.

-   First, you have the opportunity to edit your page style and header, which is called a [YAML header](https://monashdatafluency.github.io/r-rep-res/yaml-header.html), the following is an example of my own `YAML` header for this class.  

```{r, eval=FALSE}
---
title: "Lecture 1: R-Review-2023"
author: "Lifeng Ren"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: united
    toc: true
    toc_float: true
    toc_depth: 2
    highlight: tango
    df_print: paged
    mathjax: local
    self_contained: false
    number_sections: true
    fig_width: 7
    fig_height: 6
    fig_caption: true
    code_folding: hide
---
```

In the `Rmd` documentation, we have two main things:

- Markdown Document
  - Math: use the `$` sign, and the mathematical equations coding style should be the same for LaTex. 
    - For example: The probability density function of a normal distribution could be typed as: `$f(x)=\frac{1}{\sigma \sqrt{2 \pi}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$`, and `Rmd` will show the and inline output: $f(x)=\frac{1}{\sigma \sqrt{2 \pi}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$. Using `$$` instead of the single `$`, will have the following output.  
        $$
          f(x)=\frac{1}{\sigma \sqrt{2 \pi}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}
        $$
  - Hyperlink: `[This is the TEXT](This is the LINK)`. Please use the source code of this document to see the example.
  - For other Markdown coding documentation, please refer to: [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
  
- Code Chunk (mainly should be R, but can also be customized)

  The code chunk normally starts with `r`, and follows by the chunk name you defined, and then put the chunk environment after it. For example:
  
    ```{r chunkexmaple_outer, eval=FALSE}
    ``` {r chunkexmaple, eval=FALSE}
    summary(cars)
    ```  # Example 
    ```

According to the online documentation of [R Markdown](https://rmarkdown.rstudio.com/lesson-3.html):


  - Using `include = FALSE` hides both the code and its output in the final document. However, R Markdown still executes the code in the chunk, making the results accessible to subsequent chunks.
  - Setting `echo = FALSE` ensures only the code is hidden, while its output remains visible in the final document. It's particularly handy for displaying visuals without the accompanying code.
  - If you want to hide messages produced by the code, use `message = FALSE`.
  - To suppress warnings from being displayed in the final output, employ `warning = FALSE`.
  - Add a caption to images or plots using `fig.cap` = "...".
  - To skip the execution of a particular code chunk altogether, use `eval = FALSE`.
  
  - You can also use `knitr::opts_chunk$set(ANY OPTIONS ABOVE)` like below to set a global environment for all code chunks like this: 
    ```{r setup, include=TRUE, eval=FALSE}
    knitr::opts_chunk$set(echo = TRUE)
    ```

More coding details in `Rmd` should be reviewed using the source code I provided. 

## Coding Style

We are not professional programmer, but coding habitat is super important. Here are some most important parts I think you should be careful of. 

### Naming

- Please name your variables in either the following way, and being consistent:

  * `MyVariable`

  * `my_variable`

  * `my.variable`
  
- Do not start a name with numbers like: `2023badname`

- Do not include illegal characters like: `2023/%badname`

### Comments

- Remind yourself and other collaborators with comments in addition to headers.

- For example: 

    ```{r comments_example, eval=FALSE}
      
    #__________________________________________
    ##  Comments Your Code is Important----   
    #__________________________________________
    
    mean(x) # You might not need to generate comments for a simple function
    
    
    # A function to generate the Fibonacci sequence
    fibonacci <- function(n) { # but for a complex function you need more comments
      if (n <= 0) {
        return(integer(0))  # return an empty integer vector for non-positive n
      } else if (n == 1) {
        return(0)
      } else if (n == 2) {
        return(c(0, 1))
      } else {
        fib_sequence <- c(0, 1)
        for (i in 3:n) {
          next_element <- sum(tail(fib_sequence, 2))
          fib_sequence <- c(fib_sequence, next_element)
      }
      return(fib_sequence)
      }
    }

    # Print the first 10 numbers of the Fibonacci sequence
    print(fibonacci(10))
  
    ```


### Most used Built-in Symbols and Command


| Symbol | Definition                      | Example |
|--------|---------------------------------|---------|
| `=`    | Assigns within functions/datasets. | `fun(arg = value)` |
| `<-`   | Assigns values to objects/datasets.      | `obj <- 2` (shortcut: Mac: `cmd`+ `-`; Win: `ctrl`+`-`) |
| `==`   | Checks equality.                | `obj == 2` checks if `obj` equals 2. |
| `!=`   | Checks inequality.              | `obj != 2` checks if `obj` isn't 2. |
| `>`    | Greater than.                   | `obj > 2` checks if `obj` is more than 2. |
| `<`    | Less than.                      | `obj < 2` checks if `obj` is less than 2. |
| `>=`   | Greater than or equal to.       | `obj >= 2` |
| `<=`   | Less than or equal to.          | `obj <= 2` |
| `!`    | NOT (logical negation).         | `!TRUE` returns `FALSE`. |
| `&`    | AND (element-wise).             | `c(TRUE, FALSE) & c(TRUE, TRUE)` returns `TRUE, FALSE`. |
| `\|`   | OR (element-wise).              | `c(TRUE, FALSE) \| c(FALSE, FALSE)` returns `TRUE, FALSE`. |
| `&&`   | AND (first element).            | `TRUE && FALSE` returns `FALSE`. |
| `\|\|` | OR (first element).             | `TRUE \|\| FALSE` returns `TRUE`. |
| `%in%` | Tests if in a set.              | `2 %in% c(1, 2, 3)` returns `TRUE`. |

We will see them more often in the next couple sessions.

## In-class exercise

Now, let us go to `Rstudio` and play with what we have learned. 

1. Download the files

  - Create a folder in your local computer for the this class, preferably named as: `R_Review_2023`

  - Download the `lec1` folder/zip file under `R_Review_2023` from Canvas or [GitHub Repository]()

  - Open `lec1_stu.R`: Change the Script Information with your own names, date, ...

2. `print`: "Hello World"

  - In the `source` window, type the following code and select the sentence and click on `Run`
  
    ```{r helloworld}
      print("Hello World!")
    ```
  - Type the same thing in the `Console` window, and hit `enter`.
  
  - Get Help with any functions like `print`: `?print()`
  
    ```{r helloworld_help}
      ?print()
    ```
  
3. Assign a `char` value to `hello` using the following code and see the change in the `envrionemt` window. Then, print out `hello`.

    ```{r helloworld_char}
      hello <- "Hello World!"
      print(hello)
    ```
  
  
# Summary

In this section, we briefly go through Rstudio with some coding rules/conventions and get a better sense of the Rstudio Interface. After this section, make sure you are comfortable with the following points:

- Interface about Rstudio
  - How to generate a new `Rscript` and `RMarkdown` file.
- Coding Rules in R
  - How to generate the header, comments in `Rscript`
  - Basic `RMarkdown` syntax
  - Variables Naming Rules
  - Frequently used symbols


# Reference

* Section 3 of [`R for Graduate Students`](https://bookdown.org/yih_huynh/Guide-to-R-Book/rstudio-set-up.html)

* Dr. Qingxiao Li's notes for R-Review 2020

* Rodrigo Franco's notes for R-Review 2021

